Modelling Structured Societies: a Multi-relational Approach to Context Permeability

### 2015-06-20 doc/main.tex
Model description reviewed

\noindent In most simulations of complex social phenomena, agents are considered to inhabit a space in which structure is very simple, and has little resemblance with the social world it is supposed to depict, and for which the conclusions are supposed to be extrapolated. This simplicity does not come by chance, rather, it is somewhat necessary and desired by the researchers, since the problems to be approached are themselves so complex that whichever factors of complexity can be reduced, or at least postponed, are always welcome. So, as widely visible, geographical space is usually reduced to a two-dimensional grid composed by square patches, and all social relations between  agents are condensed into one abstract relation. Usually, approaches disregard the fact that in real world our relations are multiple, and, moreover, possess very diverse characteristics, including rich information about degree, intensity, frequency, methods associated, asymmetry, and so on.

To explore the addition of multiple relations and its consequences for the dissemination of phenomena in social simulations, we put forward to reduce the emphasis given on the individual interactions, depicted as a theoretical game. We accomplished this by choosing a simple, and especially, neutral game to represent those interactions. Our main concern was that the game itself would not provide biases or trends in the collective phenomenon being studied, so we chose the consensus formation game, with a straightforward majority rule to decide the outcome of each individual move.

	Studying simulation models of consensus formation can advance our understanding of real-world emergent social phenomena. Typically, we can construct models with different levels of abstraction. Simulation methods can range from data-driven paradigms to more abstract descriptions that allow us to create \textit{what-if-scenarios}. While most models are simplified descriptions of the real-world corresponding target phenomena, many constitute complex systems themselves and thus the need for tools and techniques like simulation to explore their properties. In \cite{Emmeche1997}, Emmeche makes a distinction between two types of complexity. He characterises the complexity of real systems as \textit{``ontological complexity''} and the one of models as \textit{``descriptive complexity''}. Examples of real-world target phenomena for social simulation models include for instance the joint assessments of policies or, in the context of economics and politics, \textit{the voting problem}. Herbert Simon's investigations on this problem were also one of the first stepping stones to the field of social simulation \cite{Simon1954}.

	In agent-based \textit{opinion dynamics}, agent interactions are guided by \textit{social space} abstractions. In some models, dimensionality and structure are irrelevant and any agent can interact with any other agent.
	Other approaches use an underlying structure to create agent neighbourhoods. Axelrod for instance, in its model of \textit{dissemination of culture} \cite{Axelrod1997}, represents agents in a bi-dimensional grid. Finally, in an attempt to mimic real social systems, some simulation models make use of \textit{complex network models} (see \cite{Weisbuch2004}) to create the infrastructure that guides agent interaction.

	In real-world scenarios, actors engage in a multitude of social relations different in kind and quality. Most simulation models don't explore \textit{social space} designs that take into account the differentiation between coexisting social networks. Modelling multiple concomitant relations was an idea pioneered by Peter Albin in \cite{Albin1975} but without any further development. The complex social structures that result from interacting in these different dimensions form the basis for the formation of social identity \cite{Roccas2002,Ellemers2002}.

	%******************************************
	%		Objectives
	%******************************************
	In this article, we explore the possibility of modelling opinion dynamics with multiple social networks. We look at how the properties influence the convergence to opinion consensus. Starting with this premise of 
	using multiple social networks, we present a series of models that use these networks in different ways, which create distinct emergent dynamics. We want to observe the consequences of using multiple social networks at the same time --- while maintaining the model as simple and abstract as possible, much like what happens with the current contributions on opinion dynamics present in the literature.

	In our models agents can: interact at the same time in the multiple networks (choosing partners from any network) or switch between networks (choosing only partners from their current network). While our models seem quite simple, and are quite abstract, they generate quite complex outcomes and thus the need of simulation to understand their dynamics.

	%******************************************
	%		Article Structure
	%******************************************
	\subsection{article Structure}
\noindent This article is organised as follows. In the next section, we present work related to \textit{opinion dynamics}, \textit{social space} modelling and \textit{complex network models}. In the following section, we describe our game of consensus and introduce multiple model variations, designed to study the notion of context permeability and the process of consensus formation in multiple social networks. Section \ref{sec:experimental-setup} describes the experimental set-up; the set of tool and methodologies followed to conduct our investigations. In section \ref{sec:results-discussion}, we present and discuss our results and compare the different simulation models. Finally, we conclude by making a summary of what we learned from the simulation models and point out possible future research directions.

	%******************************************
	%		Related Work
	%******************************************
	\section{Related Work}
	\label{sec:relatedwork}
\noindent In this section, we present work related to, \textit{opinion dynamics}, \textit{social space} modelling and \textit{complex network models}.	
	%******************************************
	%	Opinion Dynamics
	%******************************************
	\subsection{Opinion Dynamics and Consensus Formation}
\noindent Formal opinion dynamics models provide an understanding, if not an analysis, of opinion formation processes.  An early formulation of these models was designed to comprehend complex phenomena found empirically in groups \cite{French1956}. In particular, the work on consensus building in the context of decision-making was first studied by DeGroot \cite{Degroot74} and Lehrer \cite{Lehrer1975}. Empirical studies of opinion formation in large populations have methodological limitations, as such, computational sociology arises with a set of tools to cope with those limitations. We use simulation, in particular \textit{multi-agent simulation (MAS)}, as a methodological framework to study such phenomena in a larger scale. Most opinion dynamics simulation models are based either on binary opinions \cite{Galam1997,Antunes2009} or continuous values \cite{Deffuant2000,Deffuant2002}. In these models, agents update their opinions either under social influence or according to their own experience. For a detailed analysis over some opinion dynamics models, refer to \cite{Hegselmann2002}.

	Opinion dynamics models try to study under which circumstances consensus, or polarisation is reached. Agent-based models can have broader application outside social simulation. In distributed systems in general, consensus is a means by which processes agree on some data value that is needed during computation. Typically, this agreement is the result of a negotiation process often with the aid of a mediator. In human societies, social conventions emerge to deal with coordination and subsequently with cooperation problems \cite{Lewis1969}. These conventions are regularities of behaviour that can turn normative with time, due to there being persistent solutions to such problems. Moreover, conventions typically originate from the necessity to resolve conflict. MAS can also benefit from being capable of produce emergent conventions in coordination or cooperation problems \cite{Delgado2002}. The decentralised nature of these models of consensus is highly desirable for problems where dynamic control is needed. In these scenarios, creating conventions before hand (off-line), or developing a central control mechanism for generating them, can be a difficult and intractable task. MAS can be designed to update their values and strategies based on this local information. Consequently, the guarantee of achieving a global consensus and the efficiency of convergence towards it has been a natural concern in the design of local interaction behaviours.

	Consensus models are also investigated as means to create conventions in MAS. Shoham and Tennenholtz \cite{Shoham1994} create a bridge between economic literature and machine learning, by studying a series of models in which multiple agents are engaged in learning a particular convention. They call this process \textit{co-learning}. The complexity of these systems derives from its concurrent nature: one agent adapts to the behaviour of another agent it has encountered, and as these agents update their behaviour in a similar fashion, this results in a highly non-linear system dynamics. In their work, Shoham and Tennenholtz define the notion of \textit{stochastic social games} and compare different rules with the objective of establishing conventions in a decentralised fashion. In their stochastic games, they present one particular rule that we use in our models:

@@

	This EM was shown to coincide with \textit{Highest Cumulative Reward (HCR)}, which is a simple rule that states that ``an agent should adopt an action that has yielded the highest cumulative reward to date.'' We chose to use EM in our simulation models for its simplicity and success in the evolution of conventions. We can see that there is an overlap between opinion dynamics from social simulation and agent based learning---or co-learning, in this case.

	%******************************************
	%	Social Structure in Simulation Models
	%******************************************
	\subsection{Social Structure in Simulation Models}
\noindent One of the most widely used approaches to represent social spaces in abstract simulation models is using a  bi-dimensional grid. This approach has its origins in the first \textit{``checkerboard''} model introduced in computational social sciences by Thomas Schelling \cite{Schelling1971} and Sakoda \cite{Sakoda1971}. Another famous example that uses bi-dimensional grids as social structure is the social simulation model of dissemination of culture from Axelrod \cite{Axelrod1997}.

	Cellular automata (CA) are an example from the area of artificial life that also makes use of bi-dimensional structures to model neighbourhood interactions. While these models are idealised frameworks, the exploration of simple computational models can bring us deeper insights than models with high level descriptive complexity, which would render their analysis very difficult. The work of Flache and Hegselmann \cite{FlacheHegselmann2001} relaxes the standard CA assumptions about the regular bi-dimensional grids, by considering irregular grid structures (Voronoi diagrams). This work presents results on the robustness of some important general properties in relation to variations in the typical grid structure. We can also find contributions that incorporate both the complexity of network models and the behaviour of dynamic processes. One example of this combination is the work in \cite{Sullivan2000}, which explores a graph-based cellular automaton to study the relationship between spatial form of urban systems and the robustness of different process dynamics under spatial change. Moreover, this allows for the integration of real geographical information to construct the graph-cellular automata. This fact causes that, while the models of dynamic processes are relatively abstract, they are tested in spatial structures based on data from real complex structures and thus making a connection between purely abstract models and data-driven models.

	%******************************************
	%	Complex Network Models
	%******************************************	
	Finally, another type of social space models are the \textit{random graph/network models}. Networks with different properties can be built using network models. Each complex network or class of complex networks, presents specific topological features. These properties are typically found in---and the models inspired by---real-world network structures. One of the first complex network models is the random graph from Erd\H{o}s and Rényi \cite{Erdos1959}. Other examples include the model of preferential attachment \cite{Barabasi1999} and the small-world network models \cite{Watts1998}. One common problem of these models is the fact that once generated, the network structure is static so these models are not really suitable for the description of highly dynamic groups or communities. An example of a model that takes these dynamics into account is the model of team formation~\cite{Gaston2005}. A more in-depth review of network models and their properties is given in~\cite{Nunes2012}.

	\section{Multi-context Models}
	\label{sec:multi-context-models} 
\noindent In this section, we present our modelling approach to explore the concept of permeability between contexts. In our approach, we consider multiple social networks as a way to represent the complex social space in which an agent is inserted. In a simulation, this setting can be seen as a n-dimensional scenario, where each dimension surface represents a different social relation (see figure \ref{img:multiple-relations}). Each relation is instantiated using a social network model. The word ``context'' is used in many different senses and it is subjected to many analysis \cite{Hayes1997}. Here, we use the term \textit{social context} in a simpler and more strict way. Agents belong to distinct social contexts (neighbourhoods) in these multiple networks. The \textit{social context} of an agent is thus a set of neighbours currently available for interaction. \textit{Context Permeability} is the ability of a particular norm, strategy, opinion or trend to permeate from one social relation to another. Agents belonging to disjoint neighbourhoods in different networks foster the creation of bridges between subsets of the population that wouldn't otherwise be influenced by a consensus formed in a distant coalition or in a distinct context.

@@

	%Consensus
	We study the notion of context permeability in different simulation models in which agents interact using a simple consensus game. The society of agents has to adopt a binary \textquotedblleft option\textquotedblright according to a majority rule. We use the speed of consensus as a measure for self-organisation, exploring the relationship between different network topologies and this measure.

	%Presenting 3 models
	We partitioned the design space of our experiments using different simulation models. Each model is designed to analyse different aspects of social context permeability. In a first model \cite{Antunes2007, Antunes2010}, we study the notion of permeability by overlapping social networks. In a second model \cite{Antunes2009}, we analyse the dynamics introduced by switching between \textit{social contexts}(neighbourhoods in different networks), adding a temporal component to neighbourhoods within different social relations.

	\subsection{Simulation and Consensus Game}
	\label{sec:sim_consensus}
\noindent Our agent-based models are designed as \textit{discrete-event} models. On each simulation cycle, every agent executes a simulation step. The agents are selected in a random uniform fashion for execution. This is common practice in agent-based simulation and guarantees that there is no bias caused by the order in which the agents are executed. For each simulation model, we present a description of the individual agent behaviour. Every model we present can be seen as a binary opinion dynamics model, or a consensus game. In this game, the society of agents tries to reach an arbitrary global consensus about two possible choices. Each agent keeps track of the number of choices observed from interaction partners throughout a simulation run. In each iteration, each agent selects an available neighbour to interact with, and observes its current choice. The agent decides to switch its current choice if the observed option becomes the majority of the two possible choices. The objective of the game is to reach a global consensus, but the particular choice that gets collectively selected is irrelevant. What is important is that overall agreement is achieved. In the consensus game we consider, each agent uses the previously presented external majority (EM) rule to update its opinion value (see definition \ref{def:external-majority} in section \ref{sec:relatedwork}).

	The reason we are not really interested in which exact option gets selected is the same reason why we chose such a content-neutral game for the individual interactions. From the very beginning, our research questions were focused on the collective properties, the structure of the networks involved, and the dynamical outcomes of the simulations: thus leaving out issues related with individual motivations and desired results, either for any individual agent or the society. Granted, most applications of this research will often include such goals, for inquisition into complex social issues, including possibly policy design, in which the individual rationalities must be considered, and will have a very strong influence on the collective outcomes. We opted not to taint the simulation with such matters, leaving that very interesting approach for future work.

%	\newpage
	\subsection{Context Permeability}
\noindent This first model of context permeability \cite{Antunes2007,Antunes2010} is designed to study our first setting for permeability. In most interesting applications, an agent is immersed in a complex social world, where it engages in a multitude of relations with other agents (and even aggregated agents such as institutions or informal groups). In this model, we consider that the \textit{context permeability} is created by agents that can belong simultaneously to several relations. As an example, two agents can be simultaneously family and co-workers, and while some links might not connect two agents directly, others can contain neighbourhoods in which they are related~(either directly or through a common neighbour). Model \ref{model:context_permeability} describes the behaviour of each agent for each simulated step.
	%***************************************************
	%	Context Permeability Model Description
	%***************************************************
	\begin{algorithm}[H]
		\caption{Context Permeability}
		\label{model:context_permeability}
		\algsetup{indent=2em}
		\begin{algorithmic}
			\vspace{0.5em}
			\STATE $M$ \COMMENT{Simulation Model / Environmnet}
			\STATE $A$ \COMMENT{Current Agent in Execution}
			\\ \hrulefill

@@

	\subsection{Context Switching}
	\label{sec:models_cs}
\noindent In the context switching model \cite{Antunes2009}, agents are embedded in multiple relations represented as static social networks and they switch \textit{contexts} with a given probability $\zeta_{C_i}$ associated with context $C_i$. The switching probability value is the same for every agent in a given network.  Each agent switches from its current neighbourhood to another network (in which this agent was inactive). The agents are only active in one context at a time, and can only perform encounters with neighbours in the same context. We can think of context switching as a temporary deployment in another place, such as what happens with temporary immigration. The network topology is static; when an agent switches from one network to another, they become inactive in one network and active in their destination. 
	%***************************************************
	%	Context Switching Model Description
	%***************************************************
	\begin{algorithm}[H]
		\caption{Context Switching}
		\label{model:context_switching}
		\algsetup{indent=2em}
		\begin{algorithmic}
			\vspace{0.5em}
			\STATE $M$ \COMMENT{Simulation Model / Environmnet}
			\STATE $A$ \COMMENT{Current Agent in Execution}
			\\ \hrulefill

			\STATE\COMMENT{\textbf{Get the current network (\textit{context} is a network index)}}
			\STATE cNetwork $\leftarrow$ M.networks[A.context]
			\STATE 
			\STATE \COMMENT{\textbf{Get the neighbours for the current network}}
			\STATE neighbours $\leftarrow$ cNetwork.neighboursOf($A$) 
			\STATE \COMMENT {filter by agents active in the same network}
			\STATE neighbours $\leftarrow$ \{n $|$ n $\in$ neighbours $\wedge$ (n.context = A.context)\}
			\STATE
			\STATE\COMMENT{\textbf{Randomly select a neighbour}} 
			\STATE $r$ $\leftarrow$ randomUniform(0,neighbours.length())
			\STATE partner $\leftarrow$ neighbours[r]
			\STATE
			\STATE \COMMENT{\textbf{Update opinion and memory}}
			\STATE \COMMENT{opinions take the values \{0,1\}--so they can be used as an index}
			\STATE A.memory[partner.opinion]++
			\IF {(A.memory[partner.opinion] $>$ A.memory[A.opinion])}
			\STATE A.opinion $\leftarrow$ partner.opinion
			\ENDIF
			\STATE
			\STATE \COMMENT{\textbf{Switch to another network}}
			\STATE switchingProb $\leftarrow$ M.params.switchingProb(A.context)
			\STATE $r \leftarrow $ randomUniform(0,1)
			\IF{($r < $ switchingProb)}
			\STATE numNets $\leftarrow$ M.networks.length()
			\STATE nextNetworks $\leftarrow$ \{i $|$ (i $\in$ [0,numNets[) $\wedge$ (M.networks[i] != A.context)\}
			\STATE $r$ $\leftarrow$ randomUniform(0,nextNetworks.length())
			\STATE A.context $\leftarrow$ nextNetworks[r]
			\ENDIF
		\end{algorithmic}
	\end{algorithm}
	\noindent In this model, agents select an (active) partner from their current neighbourhood. If a partner is available, the agents update their opinion value based on EM. At the end of the interaction, the agent switches \textit{from} the current context to a different one with a probability associated with its current network.

	This model describes an abstract way to represent the time spent on each network using the switching probability $\zeta_{C_i}$. Here, the permeability between contexts is achieved using this temporal component. Context switching introduces one notion that has not been explored in the literature so far: the fact that, although some social contexts can be relatively stable if we consider short to moderate periods of time, our social peers are not always available at all times and spend different amounts of time in distinct social contexts.

>>>>

        \noindent In most simulations of complex social phenomena, agents are considered to inhabit
        a space in which structure is very simple. This space has little resemblance with the social
        world it is supposed to depict and for which conclusions are supposed to be
        extrapolated. This simplicity does not come by chance, rather, it is necessary and desired
        by the researchers: the problems to be approached are themselves so complex that whichever
        factors of complexity can be reduced (or at least postponed), the reduction is always
        welcome. So, it is common practice that geographical space is reduced to a two-dimensional
        grid, and all social relations between agents are condensed into one more or less structured
        abstract relation. Most social simulation and modelling approaches disregard the fact that
        we engage in multiple social relations. Moreover, each kind of social relation can possess
        distinctive characteristics that include: rich information such as degree of connectivity,
        centrality, trust, interactions frequency, asymmetry, and so on.

        To explore the addition of multiple relations and its consequences for the dissemination of
        phenomena in social simulations, we put forward to \textit{reduce} the emphasis given on
        agent individual interactions. We accomplished this by choosing a simple -- and especially
        neutral -- game to model those interactions. Our main concern was that the game itself would
        not provide biases or trends in the collective phenomenon being studied, so we chose the
        \textit{consensus formation game}, with a straightforward majority rule to decide the
        outcome of each individual move.

        While most models are simplified descriptions -- reductions -- of real-world phenomena, many
        constitute complex systems themselves and thus, we need techniques such as simulation to
        explore their properties. We can make a distinction between two types of complexity: the
        complexity of real systems (\textit{ontological complexity}) and the complexity of models
        (\textit{descriptive complexity})~\cite{Emmeche1997}. The levels of abstraction in a
        simulation model can then range from data-driven paradigms to more abstract descriptions
        that allow us to create \textit{what-if-scenarios}. The study of this abstract
        \textit{descriptive complexity} in simulation models is as valuable as its
        \textit{data-driven} counterpart. We may not be able to make predictions based on the direct
        application of such models; but, their study can inspire the engineering of artificial
        complex systems and reveal properties with applications beyond the explanation of observable
        phenomena. Studying consensus formation can thus advance our understanding of related
        real-world social phenomena.

        Examples of thematic real-world phenomena in social simulation models include for instance
        the joint assessments of policies or, in the context of economics and politics, \textit{the
          voting problem}. Herbert Simon's investigations on this problem were one of the first
        stepping stones to the field of social simulation \cite{Simon1954}; which then inspired
        models (like the one we present in this article) related to evolution and dissemination of
        opinions, which we call \textit{opinion dynamics}. In agent-based \textit{opinion dynamics},
        agent interactions are guided by \textit{social space} abstractions. In some models, the
        dimensionality and structure of this space is irrelevant (any agent can interact with any
        other agent). Other models use an underlying artefact that structures agent
        neighbourhoods. Axelrod for instance, represents agent neighbourhoods as a bi-dimensional
        grid in its model of \textit{dissemination of culture} \cite{Axelrod1997}. In an attempt to
        mimic real social systems, one can also make use of \textit{complex network models} to
        create the infrastructure that guides agent interaction (see \cite{Weisbuch2004} for an
        example).

	In real-world scenarios, actors engage in a multitude of social relations different in kind
        and quality. Most simulation models don't explore \textit{social space} designs that take
        into account the differentiation between coexisting social networks. Modelling multiple
        coexisting relations was an idea pioneered by Peter Albin in \cite{Albin1975} but without
        further development. The process of interacting in these different complex social dimensions
        can be seen as the basis for the formation of our social identity
        \cite{Roccas2002,Ellemers2002}.

	%******************************************
	%		Objectives
	%******************************************
	In this article, we explore the of modelling opinion dynamics with multiple social
        networks. We look at how the properties of different network models influence the
        convergence to opinion consensus. We present a series of models that use these networks in
        different ways: which create distinct emergent dynamics. We want to study the consequences
        of using multiple social networks at the same time while maintaining the interaction model
        as simple and abstract as possible (following the methodology in most opinion dynamics
        literature). We present two models where agents can: interact at the same time in the
        multiple networks (choosing partners from any network); or switch between networks (choosing
        only partners from their current network).

	%******************************************
	%		Article Structure
	%******************************************
	\subsection{Article Structure}
        \noindent This article is organised as follows. In the next section, we present the work
        related to \textit{opinion dynamics}, \textit{social space} modelling and \textit{complex
          network models}. In the following section, we describe our game of consensus and introduce
        our multiple model variations. These are designed to study the notion of \textit{context
          permeability} and consensus formation in multiple social networks. Section
        \ref{sec:experimental-setup} describes the experimental setup; the set of tool and
        methodologies followed to conduct our investigations. In section
        \ref{sec:results-discussion}, we present and discuss our results and compare the different
        simulation models. Finally, we summarise what we learned from our experiments and point out
        future research directions.

	%******************************************
	%		Related Work
	%******************************************
	\section{Related Work}
	\label{sec:relatedwork}
        \noindent In this section, we present work related to \textit{opinion dynamics},
        \textit{social space} modelling, and \textit{complex network models}.
	%******************************************
	%	Opinion Dynamics
	%******************************************
	\subsection{Opinion Dynamics and Consensus Formation}
        \noindent Formal opinion dynamics models provide an understanding, if not an analysis, of
        opinion formation processes. An early formulation of such models was created as a way to
        understand complex phenomena found empirically in groups \cite{French1956}. The work on
        consensus building (in the context of decision-making) was first studied by DeGroot
        \cite{Degroot74} and Lehrer \cite{Lehrer1975}. Empirical studies of opinion formation in
        large populations have methodological limitations. Computational sociology arises with a set
        of tools -- simulation models in particular -- to cope with such limitations. We use
        \textit{multi-agent simulation (MAS)} as a methodological framework to study such social
        phenomena in a larger scale. Most opinion dynamics models make use of binary opinion values
        \cite{Galam1997,Antunes2009}, or continuous values \cite{Deffuant2000,Deffuant2002}. For a
        detailed analysis over some opinion dynamics models, refer to \cite{Hegselmann2002}.

	Opinion dynamics models allow us to discover under which circumstances a population of
        agents reaches consensus or polarisation. Agent-based models can have broader application
        outside social simulation though. In computational distributed systems, consensus is a means
        by which processes agree on some data~value needed during computation. Typically, this
        agreement is the result of a negotiation process (often with the aid of a mediator). In
        human societies, social conventions emerge to deal with coordination and subsequently with
        cooperation problems \cite{Lewis1969}. These conventions are regularities of behaviour that
        can turn normative if they come to be persistent solutions to recurrent problems. MAS are
        also capable of producing emergent conventions in coordination or cooperation problems
        \cite{Delgado2002}. The decentralised nature of these computational models of consensus is
        highly desirable for dynamic control problems. In these scenarios, creating conventions
        before hand (off-line), or developing a central control mechanism for generating them, can
        be a difficult and intractable task: either due to the uncertainty and complexity associated
        with the environment, or due to the system scale and heterogeneity (which makes it very
        difficult to handcraft each component).

	Consensus models are also investigated as means to create conventions in MAS. Shoham and
        Tennenholtz create a bridge between economic literature and machine learning by studying a
        series of models in which multiple agents are engaged in learning a particular convention
        \cite{Shoham1994}. They call this process \textit{co-learning}. The complexity of these
        systems comes from its concurrent nature: one agent adapts to the behaviour of another
        agents it has encountered, these other agents update their behaviour in a similar fashion,
        which results in a highly non-linear system dynamics. In their work, Shoham and Tennenholtz
        define the notion of \textit{stochastic social games} and compare different rules with the
        objective of establishing conventions in a decentralised fashion. In their stochastic games,
        they present one particular rule that we use in our models:

@@

	This EM rule was shown to coincide with \textit{Highest Cumulative Reward (HCR)}, which is a
        simple rule that states that ``an agent should adopt an action that has yielded the highest
        cumulative reward to date.'' We use the EM rule in our simulation models for its simplicity
        and success in the evolution of conventions.

	%******************************************
	%	Social Structure in Simulation Models
	%******************************************
	\subsection{Social Structure in Simulation Models}
        \noindent The usage of a bi-dimensional grid to represent abstract social spaces in
        simulation models, is one of the most widely used approaches in the agent-based simulation
        and modelling literature. This has its origins in the \textit{``checkerboard''} model
        introduced in computational social sciences by Schelling and Sakoda
        \cite{Schelling1971,Sakoda1971}. Another famous example that uses bi-dimensional grids as
        social structure is the social simulation model of dissemination of culture from Axelrod
        \cite{Axelrod1997}.

	Cellular automata (CA) are an example from the area of \textit{artificial life} that also
        makes use of bi-dimensional structures to model neighbourhood interactions. While these
        models are idealised frameworks, their exploration can bring us deeper insights than models
        with high level of descriptive complexity (which would render their analysis very
        difficult). The work of Flache and Hegselmann relaxes the standard CA assumptions about the
        regular bi-dimensional grids by considering irregular grid structures (Voronoi diagrams)
        \cite{FlacheHegselmann2001}. They present results on the robustness of some important
        general properties in relation to variations in the typical grid structure. There are also
        contributions that incorporate both the complexity of network models and the behaviour of
        dynamic processes. One example is the work in \cite{Sullivan2000}, which explores a
        graph-based cellular automaton to study the relationship between spatial forms of urban
        systems and the robustness of different process dynamics under spatial change. Moreover,
        this shows how we can use real geographical information to construct graph-based cellular
        automata (thus making a connection between purely abstract models and data-driven models).

	%******************************************
	%	Complex Network Models
	%******************************************	
	Finally, another type of social space models are the \textit{random graph/network
          models}. Each complex network -- or \textit{class} of complex networks -- captures
        specific topological properties. These properties are found in real-world network
        structures. One of the first complex network models is the random graph from Erd\H{o}s and
        Rényi \cite{Erdos1959}. Other examples include the model of preferential attachment
        \cite{Barabasi1999}, and the small-world networks \cite{Watts1998}. One common issue of
        these models is the fact that once generated, the network structure is static. The models
        are not really suitable for the description of highly dynamic groups or communities. An
        example of a model that takes such dynamics into account is the model of team
        formation in~\cite{Gaston2005}. A more in-depth review of network models (and their properties)
        is given in~\cite{Nunes2012}.

	\section{Multi-context Models}
	\label{sec:multi-context-models} 
        \noindent In this section, we present our modelling approach to explore the concept of
        permeability between contexts. We use multiple social networks to represent the complex
        social space in which an agent is inserted. In a simulation model, this setting can be seen
        as a n-dimensional scenario where each dimension contains a network that represents a
        different social relation (see figure \ref{img:multiple-relations}). The word ``context'' is
        used in many different senses and it is by itself subjected to many analysis
        \cite{Hayes1997}. Here, we use the term \textit{social context} in a simpler and more strict
        way. Agents belong to distinct social contexts which are their neighbourhoods in these
        multiple networks. The \textit{social context} of an agent is thus a set of neighbours
        currently available for interaction at a particular network or set of
        networks. \textit{Context permeability} is the ability of a particular norm, strategy,
        opinion, or trend, to permeate from one social relation to another. Agents belonging to
        disjoint neighbourhoods in different networks serve as bridges between subsets of the
        population that wouldn't otherwise be influenced by a consensus formed in a distant
        coalition.

@@

	%Consensus
	We study the notion of context permeability in different simulation models in which agents
        interact using a simple consensus game. The society of agents has to adopt a binary opinion
        value according to a majority rule. We use the speed of consensus as a measure for
        self-organisation and explore the relationship between different network topologies and this
        measure.

	%Presenting 2 models
	We partitioned the design space of our experiments using different simulation models. Each
        model is designed to analyse different aspects of context permeability. In a first model
        \cite{Antunes2007, Antunes2010}, we study the notion of permeability by overlapping social
        networks. In a second model \cite{Antunes2009}, we analyse the dynamics introduced by
        switching between \textit{social contexts}(neighbourhoods in different networks). This adds
        a temporal component that changes the dynamics of the game: agents can only perform
        encounters with neighbours that are active in their current context. (It also adds the
        chance for some agents to be isolated during the simulation.)

	\subsection{Simulation and Consensus Game}
	\label{sec:sim_consensus}
        \noindent Our agent-based models are designed as \textit{discrete-event} models. On each
        simulation cycle, every agent executes a simulation step. The agents are selected in a
        random uniform fashion for execution. This is common practice in agent-based simulation and
        guarantees that there is no bias caused by the order in which the agents are executed. For
        each simulation model, we present a description of the individual agent behaviour. Every
        model we present can be seen as a binary opinion dynamics model, or a consensus game. In
        this game, the society of agents tries to reach an arbitrary global consensus about two
        possible choices or opinion values. During a simulation run, each agent keeps track of the
        number of each opinion value of their interaction partners. In each iteration, each agent
        selects an available neighbour to interact with and observes its current opinion value. The
        agent decides to switch its current opinion choice if the observed opinion value becomes the
        majority of the two possible choices. The goal of the game is to reach a global consensus,
        but the particular choice that gets collectively selected is irrelevant. What is important
        is that overall agreement is achieved. In the consensus game we consider, each agent uses
        the previously presented external majority (EM) rule to update its opinion value (see
        definition \ref{def:external-majority} in section \ref{sec:relatedwork}).

	The reason we are not interested in which exact option gets selected is the same reason why
        we chose such a content-neutral game for the individual interactions. From the very
        beginning, our research questions were focused on the collective properties, the structure
        of the networks involved, and the dynamical outcomes of the simulations: thus leaving out
        issues related with individual motivations and desired results (either for any individual
        agent or the society). Granted, most applications of this research will often include such
        goals, for inquisition into complex social issues, including possibly policy design, in
        which the individual rationality must be considered, and will have a very strong influence
        on the collective outcomes.

	\subsection{Context Permeability}
        \noindent The first model of context permeability \cite{Antunes2007,Antunes2010} is designed
        to study the setting where an agent is immersed in a complex social world: where it engages
        in a multitude of relations with other agents. In this model, we consider that the
        \textit{context permeability} is created by agents that can belong \textit{simultaneously}
        to several relations. As an example, two agents can be simultaneously family and
        co-workers. While some links might not connect two agents directly, others can contain
        neighbourhoods in which they are related~(either directly or through a common
        neighbour). Model \ref{model:context_permeability} describes the behaviour of each agent for
        each simulated step.
	%***************************************************
	%	Context Permeability Model Description
	%***************************************************
	\begin{algorithm}[H]
		\caption{Context Permeability}
		\label{model:context_permeability}
		\algsetup{indent=2em}
		\begin{algorithmic}
			\vspace{0.5em}
			\STATE $M$ \COMMENT{Simulation Model / Environment}
			\STATE $A$ \COMMENT{Current Agent in Execution}
			\\ \hrulefill

@@

	\subsection{Context Switching}
	\label{sec:models_cs}
        \noindent In the context switching model \cite{Antunes2009}, agents are embedded in multiple
        relations represented as static social networks and they switch \textit{contexts} with a
        probability $\zeta_{C_i}$ associated with context $C_i$. The switching probability is the
        same for every agent in a given network.  Each agent switches from its current neighbourhood
        to another network. The agents are only active in one context at a time, and can only
        perform encounters with neighbours in the same context. We can think of context switching as
        a temporary deployment in another place, such as what happens with temporary
        immigration. The network topology is static; when an agent switches from one network to
        another, they become inactive in one network and active in their destination.

	%***************************************************
	%	Context Switching Model Description
	%***************************************************
	\begin{algorithm}[H]
		\caption{Context Switching}
		\label{model:context_switching}
		\algsetup{indent=2em}
		\begin{algorithmic}
			\vspace{0.5em}
			\STATE $M$ \COMMENT{Simulation Model / Environment}
			\STATE $A$ \COMMENT{Current Agent in Execution}
			\\ \hrulefill

			\STATE\COMMENT{\textbf{Get the current network (\textit{context} is a network index)}}
			\STATE cNetwork $\leftarrow$ M.networks[A.context]
			\STATE 
			\STATE \COMMENT{\textbf{Get the neighbours for the current network}}
			\STATE neighbours $\leftarrow$ cNetwork.neighboursOf($A$) 
			\STATE \COMMENT {filter by agents active in the same network}
			\STATE neighbours $\leftarrow$ \{n $|$ n $\in$ neighbours $\wedge$ (n.context = A.context)\}
			\STATE
			\STATE\COMMENT{\textbf{Randomly select a neighbour}} 
			\STATE $r$ $\leftarrow$ randomUniform(0,neighbours.length())
			\STATE partner $\leftarrow$ neighbours[r]
			\STATE
			\STATE \COMMENT{\textbf{Update opinion and memory}}
			\STATE \COMMENT{opinions take the values \{0,1\}--so they can be used as an index}
			\STATE A.memory[partner.opinion]++
			\IF {(A.memory[partner.opinion] $>$ A.memory[A.opinion])}
			\STATE A.opinion $\leftarrow$ partner.opinion
			\ENDIF
			\STATE
			\STATE \COMMENT{\textbf{Switch to another network}}
			\STATE switchingProb $\leftarrow$ M.params.switchingProb(A.context)
			\STATE $r \leftarrow $ randomUniform(0,1)
			\IF{($r < $ switchingProb)}
			\STATE numNets $\leftarrow$ M.networks.length()
			\STATE nextNetworks $\leftarrow$ \{i $|$ (i $\in$ [0,numNets[) $\wedge$ (M.networks[i] != A.context)\}
			\STATE $r$ $\leftarrow$ randomUniform(0,nextNetworks.length())
			\STATE A.context $\leftarrow$ nextNetworks[r]
			\ENDIF
		\end{algorithmic}
	\end{algorithm}
	\noindent In this model, agents select an (active) partner from their current
        neighbourhood. If a partner is available, the agents update their opinion value based on
        EM. At the end of the interaction, the agent switches \textit{from} the current context to a
        different one with a probability associated with its current context.

	This model describes an abstract way to represent the time spent on each network using the
        switching probability $\zeta_{C_i}$. Here, the permeability between contexts is achieved
        using this temporal component. Context switching introduces a notion that has not been
        explored in the literature so far: the fact that, although some social contexts can be
        relatively stable, our social peers are not always available for interaction and spend
        different amounts of time in distinct social contexts.
